---
title: "Using sctransform in Seurat"
author: "Christoph Hafemeister & Rahul Satija"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{Using sctransform in Seurat}  
  %\VignetteEngine{knitr::rmarkdown}  
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library('Matrix')
library('ggplot2')
library('reshape2')
library('sctransform')
library('knitr')
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  digits = 2,
  fig.width=8, fig.height=5, dpi=100, out.width = '70%'
)
library('Seurat')
#old_theme <- theme_set(theme_classic(base_size=8))
```

This vignette shows how to use the sctransform wrapper in Seurat.
Install sctransform and Seurat v3
```{r eval=FALSE}
devtools::install_github(repo = 'ChristophH/sctransform', ref = 'develop')
devtools::install_github(repo = 'satijalab/seurat', ref = 'release/3.0')
library(Seurat)
library(sctransform)
```
Load data and create Seurat object
```{r load_data, warning=FALSE, message=FALSE, cache = T}
pbmc_data <- Read10X(data.dir = "~/Downloads/pbmc3k_filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc_data)
```
Apply sctransform normalization
```{r apply_sct, warning=FALSE, message=FALSE, cache = T}
# Note that this single command replaces NormalizeData, ScaleData, and FindVariableFeatures.
# Transformed data will be available in the SCT assay, which is set as the default after running sctransform
pbmc <- SCTransform(object = pbmc, verbose = FALSE)
```
Perform dimensionality reduction by PCA and UMAP embedding
```{r pca, fig.width=5, fig.height=5, cache = T}
# These are now standard steps in the Seurat workflow for visualization and clustering
pbmc <- RunPCA(object = pbmc, verbose = FALSE)
pbmc <- RunUMAP(object = pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindNeighbors(object = pbmc, dims = 1:20, verbose = FALSE)
pbmc <- FindClusters(object = pbmc, verbose = FALSE)
DimPlot(object = pbmc, label = TRUE) + NoLegend()
```
Users can individually annotate clusters based on canonical markers. However, the sctransform normalization reveals sharper biological distinctions compared to the [standard Seurat workflow](https://satijalab.org/seurat/pbmc3k_tutorial.html), in a few ways:
 * Clear separation of three CD8 T cell populations (naive, memory, effector), based on CD8A, GZMK, CCL5, GZMK expression
 * Clear separation of three CD4 T cell populations (naive, memory, IFN-activated) based on S100A4, CCR7, IL32, and ISG15 
 * Additional developmental sub-structure in B cell cluster, based on TCL1A, FCER2
 * Additional separation of NK cells into CD56dim vs. bright clusters, based on XCL1 and FCGR3A 
```{r fplot, fig.width = 10, fig.height=10, cache = F}
# These are now standard steps in the Seurat workflow for visualization and clustering
FeaturePlot(object = pbmc, features = c("CD8A","GZMK","CCL5","S100A4"), pt.size = 0.3)
FeaturePlot(object = pbmc, features = c("S100A4","CCR7","CD4","ISG15"), pt.size = 0.3)
FeaturePlot(object = pbmc, features = c("TCL1A","FCER2","XCL1","FCGR3A"), pt.size = 0.3)
```
